---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r}
activity <-read.csv("activity.csv")
head(activity)
summary(activity)
```

The dataframe "activity" looks OK as it is, so no further preprocessing is necessary

Let's have a quick look at the data as it is:

```{r}
plot(activity)
```


## What is mean total number of steps taken per day?

Group data by date and sum the number of steps taken each day.
Show a histogram of the number of total steps taken each day and calculate the mean and median of the total number of steps taken each day:

```{r}
suppressMessages(library(dplyr))
by_date <- group_by(activity,date)
by_date <- summarise(by_date, steps=sum(steps))
head(by_date)
hist(by_date$steps, breaks = 20, col = "lightblue", main = "Steps per day Histogram", xlab = "Steps per day")
summary(by_date)
```


## What is the average daily activity pattern?

ake a time series plot of the average of the number of steps taken each 5 minute interval, averaged over all days, and show the 5 minute interval that contains the maximum number of steps on average across all the days in the dataset

```{r}
by_interval <- group_by(activity, interval)
by_interval <- summarise(by_interval, steps = mean(steps, na.rm = TRUE))
head(by_interval)

plot(by_interval, type = "l")
```

Interval of maximum steps averaged across all days:

```{r}
filter(by_interval, steps == max(steps))
```


## Imputing missing values

Number of missing values in this file : `r sum(is.na(activity$steps)) ` (this is calculated with R inline code sum(is.na(activity$steps)))

The NA values will be replaced by the mean number of steps per day in the same interval.  This mean value has already been calculated in the dataframe "by_interval" which is now used to find the imputed value for the NA's.

The dataset with the filled in NA's is called "na_imputed"

```{r}
na_imputed <- activity
na_imputed$steps <- ifelse(is.na(na_imputed$steps), by_interval$steps[match(na_imputed$interval, by_interval$interval)], na_imputed$steps)
head(na_imputed)
```

Histogram of the total number of steps taken each day, now with NA's imputed (first we have to regroup the imputed dataset by day):

```{r}
by_date_imp <- group_by(na_imputed,date)
by_date_imp <- summarise(by_date_imp, steps=sum(steps))
head(by_date_imp)
hist(by_date_imp$steps, breaks = 20, col = "lightgreen", main = "Steps per day Histogram with imputed NA values", xlab = "Steps per day")
```

The mean and median are given in the summary:

```{r}
summary(by_date_imp)
```

The impact of imputing missing data on the estimates of the total daily number of steps can be shown comparing the mean, median and histograms of both datasets:

The mean of the raw data is `r mean(by_date$steps,na.rm=TRUE)` while the mean of the imputed dataset is `r mean(by_date_imp$steps)`

The median of the raw data is `r median(by_date$steps,na.rm=TRUE)` and the median of the imputed dataset is  `r median(by_date_imp$steps)`


## Are there differences in activity patterns between weekdays and weekends?

```{r}
library(ggplot2)
wkday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
na_imputed$daytype <- as.factor(ifelse(is.element(weekdays(as.Date(na_imputed$date)), wkday), "Weekday", "Weekend"))
by_interval_and_daytype <- group_by(na_imputed, daytype, interval)
by_interval_and_daytype <- summarise(by_interval_and_daytype, steps = mean(steps))
head(by_interval_and_daytype)
qplot(interval, steps, data=by_interval_and_daytype, geom ="line", colour = daytype)
```

It looks as if activities start earlier during weekdays and reach a higher peak in the morning, while overal activity during daytime seems to be higher on weekends.

